Measuring the double-drift illusion with hand tracking

### 2021-02-25 Manuscript.Rmd
edits to manuscript

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background [@Gurnsey2012; @Kwon2015; @Cavanagh2019]. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position. If the gabor itself is not moving [@DeValois1991], the shift saturates quickly [40 _ms_: @Chung2007; 100 _ms_: @Arnold2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location up to several times the size of the gabor. The strength of the illusion, or how much the perceived position shifts from the real position over time, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015]. In our first experiment, participants track their percept with a stylus as the gabor moves back and forth on its trajectory (Fig. 1) with 4 internal speeds. We analyze the angle of the tracking to determine if manual response is susceptible to the illusion.

At some point, the illusion stops accumulating. The position may just remain at some fixed offset following a path parallel to the true path or the position may reset back toward its physical location and then accumulation begins again (Fig. 4, left) or it saturates. This long duration before resetting or saturation is well beyond the integration period of other visual effects and into the range of short-term memory. Given that both true and illusory positions are available, it could be that instead of some integration period, it is the interplay between the two position signals that leads to spontaneous resets. Alternatively, there is a simple periodicity or a stochastic process that underlies the changes in perception. Recent work has shown that independent of the stimulus, movement-related attention may lead to resets [@Nakayama2020]. This means that participants' responses to the stimulus may change the illusion. Here we will first test if real-time as well as delayed, manual tracking of the gabor's position, prevents or decreases the illusory shift in position.

The ability to record trajectories the perceived position will then allow us to study spontaneous resets – the adjustment of perceived location when the illusion has deviated too much or too long from the physical location. In particular, if a conflict between two location signals [@Lisi2015] leads to resets this would predict a set distance between the real and perceived position where resets occur. Under this hypothesis, the timing of spontaneous resets would depend on the speed of internal motion (given constant external motion). On the other hand, if spontaeous resets are the result of a process that integrates over time, resets would occur after a certain amount of time, but not be dependent on internal motion. Investigating the time and location of resets in the recorded trajectories, allows us to distinguish between a spatial limit to the illusion or a temporal limit. This in turn will inform us on the processes underlying the perception of positions of moving objects.

# Online tracking experiment

@@

For reporting degrees of visual angle that stimuli spanned ( _dva_ ), we assumed the participants to be close to the setup, with their eyes at the height of the monitor. While there was no requirement to use the forehead rest, this is a reasonable assumption, since stimuli are only visible when relatively close to the setup. Nevertless, reported _dva's_ are estimates of the size of stimuli.

@@

The motion of the envelope was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that the stimulus would move from one end to the other of its path ( a motion we call a 'pass' ) in 2, 3 or 4 _s_, corresponding to speeds of ~6.75, ~4.5 and ~3.75 _cm/s_ ( or roughly 7.0, 4.7 or 3.5 _dva/s_ ). For simplicity, and since the distance moved is always the same, these speeds will be refered to by the duration of a single pass (2, 3 or 4 _s_).

@@

In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass for each of the 5 internal motion speeds (normalized for the external motion). The average angle of the tracked path is given in degrees deviation from straight forward. It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ). With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001). This makes it clear that continuous tracking does not prevent the illusion from occurring. We will use heading angle as a measure of the strength of the illusion.

We tried detecting resets, but none could be unambiguously identified (see Fig 2). It is possible that resets were perceived but were too quick to respond to in the continuous tracking task. Alternatively, tracking may have stabilized the illusion, preventing resets from occurring.

```{r fig2, echo=FALSE, message=FALSE, fig.width=4.5, fig.height=6, fig.cap="\\label{fig:fig2}**Figure 2. Heading and Trajectories in Experiment 1.** **Top:** Average heading across the middle 1 _s_ of each pass. Distance from origin is time, angle is angle. In darker blue lines the average heading across participants, in light blue a polar heatmap giving an impression of the distribution of heading. **Bottom:** All individual trajectores. _Red_: Highlighted trajectories from conditions with non-zero internal speed, where at least some heading samples fell in the 95% confidence interval for that participants' heading distribution in zero internal speed condition. This would occur during a reset, but only very few trajectories have this feature, and none contain clear resets."}
plotBoundedTracking()
```

@@

```{r fig3, echo=FALSE, message=FALSE, fig.width=5, fig.height=5, fig.cap="\\label{fig:fig3}**Figure 3. Illusion strength.** The strength of the illusion in degrees angle deviation from the physical path is plotted over external motion speed ( _dva/s_ ). In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for individual participants. The gray lines are the predicted illusion strengths based on the model from Cavanagh & Tse (2019) for the internal motion speeds used here. *the second panel only for testing*"}
plotStrengthComparison()
```

# Delayed retracing experiment

In a second experiment, aimed at collecting resets, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus, a method used by Nakayama & Holcombe [-@Nakayama2020]. This allows participants more time to carefully reproduce the path. We use the recorded trajectories to assess points where resets occurred. The reset points can then be used to determine if resets are time-limited or space-limited.

```{r fig4b, echo=F, fig.height=3.2, fig.width=5.62, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 4: Resets.** _Left_: At some point, the perceived position will stop drifting further and further away from its actual path. For some people this takes the form of a ‘jump’ back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, and there are some in-between versions. _Top right_: if these resets are due to a spatial limit on the size of the illusory position shift, the points of the resets should have a constant X coordinate. _Bottom right_: if they occur after some period of time, they should have a constant Y coordinate."}
library(magick)
Fig4b <- image_read_svg('doc/methods_fig_2.svg')
par(mai=c(0,0,0,0))
plot(Fig4b)
```

@@

The setup was the same as used in the online tracking experiment, except that we now also used a small keypad, for responses (see Fig 1, right). Stimuli were identical to those used in the online tracking experiment, but now always started at the near end of the workspace and moved away from the participants. The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_ and had an internal drift of 2, 3 or 4 _cps_.

@@

Re-tracing the perceived paths should allow capturing spontaneous resets of the illusion, and the depence on the strength of the illusion could be checked with the data from the line-orienting task.

@@

We use the same heading measure here as in experiment 1. However, the individual trajectories reveal more points that are potential spontaneous resets that we try to extract. Using a smooth spline interpolated version of the trajectory we detect left-right direction changes. Since the initial movement is often jittery, we only take direction changes further than ~0.39 cm away from the starting point. The first (if any) of the remaining direction changes, are then taken as reset points. By averaging the x and y coordinates of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling. The average line orientation is also calculated for each of those 54 combinations of partcipants and stimulus conditions.

@@

In Figure 4, for each of the 6 combinations of internal speed (2, 3 or 4 _cps_) and external speed (3 or 4 _s_ stimulus pass) we show the data for one example participant, as well as the heading distribution for all participants. While the heading distributions don't show resets, for many of the participants, we can see that the re-traced paths do show resets, many of which are detected by our algorithm.

```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig4}**Figure 4. Example Trajectories and Heading in Experiment 2.** For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show one example participant's trajectories and reset points, as well as the heading distribution."}
plotExampleData()
```

@@

```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5. Measures of Illusion Strength** Line angle responses and reset point direction should both measure illusion strength. **Left:** Line Angle response as a function of internal and external speed, averaged across 9 participants in each of the 6 conditions. Shaded areas denote 95% confidence intervals. **Right:** Angles of samples halfway to a reset over line angle response. Each dot corresponds to one participants average in one of the 6 conditions. Diagonal line indicates a weighted regression (weighted by one over the square of the line angle response to counter heteroscedasticity), and the shaded area shows a 95% confidence interval for the regression."}
plotValidity()
```

As specified above, we also take the angle of the sample on each re-tracing trajectory that is halfway between the origin and the detected reset, as an alternative measure of illusion strength. If both the angle half-way to reset and line angle responses are good measures of illusion strength, they should be related. As can be seen from the data in Figure 5 (right), there might be a relation between the two variables, but there is also heteroscadasticity. We counter this by doing a weighted regression, with the weights set to one over the squared line angle ($\frac{1}{\alpha^2}$). Reported line angles predicted reset point direction (slope=0.71; _t_(49)=4.54, _p_<.001; $R^2$=.282; _F_(1,49)=20.6), verifying that the two measures - while imperfect - are likely both indicators of double-drift illusion strength. From here on, we will only use the angle of samples half-way to detected reset points as measure of illusion strength.

>>>>

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background [@Gurnsey2012; @Kwon2015; @Cavanagh2019]. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position. If the gabor itself is not moving [@DeValois1991], the shift saturates quickly [40 _ms_: @Chung2007; 100 _ms_: @Arnold2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues well beyond 100 ms. The perceived position conintues to move away from its physical position up to several times the size of the gabor. The strength of the illusion, or the deviation of perceived path of the Gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015]; @Cavanagh2019. In our first experiment, participants track their percept with a stylus as the gabor moves back and forth on its trajectory (Fig. 1) with 5 speeds of the internal motion. We analyze the angle of the tracking to determine if manual response is susceptible to the illusion.

In our second experiment, we look at the spatial and temporal limits of the illusion. At some point, the illusion stops accumulating. The accumulation can be reset by introducing a temporal break [@Lisi2015] or by distraction attention [@Nakayama2020], but it may also reset spontaneously once the accumulation has gone on too long or too far. Informally, some observers have reported that the Gabor’s path may saturate so that it  remains parallel to, but offset from the true path. Others report that the position resets back toward its physical location whereupon accumulation begins again. The ability to record trajectories the perceived position will then allow us to study spontaneous reset outcomes.

# Experiment 1: Online tracking

@@

For reporting degrees of visual angle that stimuli spanned ( _dva_ ), we assumed the participants to be close to the setup, with their eyes at the height of the monitor. While there was no requirement to use the forehead rest, this is a reasonable assumption, since stimuli are only visible when relatively close to the setup. Nevertheless, reported _dva's_ are estimates of the size of stimuli.

@@

The motion of the envelope was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that the stimulus would move from one end to the other of its path ( a motion we call a 'pass' ) in 2 _s_, corresponding to a speed of ~6.75 _cm/s_ ( or roughly 7.0 _dva/s_ ).

@@

In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass for each of the 5 internal motion speeds. The average angle of the tracked path is given in degrees deviation from straight forward. It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ). With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001). This makes it clear that continuous tracking does not prevent the illusion from occurring. We will use heading angle as a measure of the strength of the illusion.

```{r fig2, echo=FALSE, message=FALSE, fig.width=4.5, fig.height=6, fig.cap="\\label{fig:fig2}**Figure 2. Heading and Trajectories in Experiment 1.** **Top:** Average heading across the middle 1 _s_ of each pass. Distance from origin is time, angle is angle. In darker blue lines the average heading across participants, in light blue a polar heatmap giving an impression of the distribution of heading. **Bottom:** All individual trajectores. Note that there are only a few obvious discontinuities (highlighted in red). In general, the traced paths showed rough linearity."}
plotBoundedTracking()
```

@@

```{r fig3, echo=FALSE, message=FALSE, fig.width=5, fig.height=5, fig.cap="\\label{fig:fig3}**Figure 3. Illusion strength.** The strength of the illusion in degrees angle deviation from the physical path is plotted over external motion speed ( _dva/s_ ). In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for individual participants. The gray lines are the predicted illusion strengths based on the model from Cavanagh & Tse (2019) for the internal motion speeds used here."}
plotStrengthComparison()
```

# Experiment 2: Delayed retracing

The durations of each pass in the online tracking experiment was 2 seconds but the manual traces mostly showed a linear trajectory without saturation or reset of the types shown in Figure 4 on the left below. The continuation of the accumulation may be a result of the active tracking – either the extra attention required to track increased the extent of the accumulation or possibly the temporal demands of the tracking caused the participants to average their traces over any breaks. To address this, we switched to an offline recording of the perceived path.

```{r fig4b, echo=F, fig.height=3.2, fig.width=5.62, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 4: Resets.** _Left_: At some point, the perceived position will stop moving further away from its actual path. For some people this takes the form of a ‘jump’ back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combnation of the two. _Top right_: if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate. _Bottom right_: if they occur instead after some period of time, they should have a constant Y coordinate."}
library(magick)
Fig4b <- image_read_svg('doc/methods_fig_2.svg')
par(mai=c(0,0,0,0))
plot(Fig4b)
```

We were interested, in particular, in whether the limitation of the accumulation would be set by space or time. Since both the true and the illusory position appear to be available in the visual system (the true position drives saccades whereas the illusory position drives perception, Lisi & Cavanagh; -@Lisi2015), it is possible that resets occur once the distance between the real and perceived positions exceeds some limit. We can speculate that this spatial limit would be related to the positional uncertainty of the Gabor. When the position information is reliable, the resets would occur with very little illusory deviation, keeping the perceived path close to the true path. When the positional uncertainty is high, in the periphery, with the Gabor’s mean luminance matched to the background [@Gurnsey2012; @Cavanagh2019], the spatial offset could be quite large before exceeding the range of positional uncertainty around the gabor’s true position. Under this hypothesis, the timing of spontaneous resets will depend on the spatial offset from the true path which will be directly proportional to the internal speed (for a constant external motion). On the other hand, if spontaneous resets are the result of the temporal limitations of the integration process, resets would occur after a certain amount of time, independently of the internal speed and the spatial offset it creates. Investigating the time and location of resets in the recorded trajectories, allows us to distinguish between a spatial limit to the illusion and a temporal limit. This in turn will inform us of the processes underlying the perception of the position of moving objects.

To do so, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus, a method used by Nakayama & Holcombe [-@Nakayama2020]. This allows participants more time to carefully reproduce the path. We use the recorded trajectories to assess points where resets occurred, whether they were a discontinuity in orientation reflecting a saturation (hit the wall reset, Fig. 4) or a discontinuity in position reflecting a jump back to the real location. These reset points can then be used to determine if resets are time-limited or space-limited.

@@

The setup was the same as used in the online tracking experiment, except that we now also used a small keypad, for responses (see Fig 1, right). Stimuli were identical to those used in the online tracking experiment, but now always started at the near end of the workspace and moved away from the participants. The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_ ( corresponding to speeds of ~4.5 and ~3.75 _cm/s_ or ~4.7 or ~3.5 _dva/s_ ) and had an internal drift of 2, 3 or 4 _cps_ .

@@

Re-tracing the perceived paths should allow capturing spontaneous resets of the illusion, and the dependence on the strength of the illusion could be checked with the data from the line-orienting task.

@@

The individual trajectories reveal some points that are potential spontaneous resets that we try to extract. Using a smooth spline interpolated version of the trajectory we detect left-right direction changes. Since the initial movement is often jittery, we only take direction changes further than ~0.39 cm ( ~0.23 _dva_ ) away from the starting point. The first (if any) of the remaining direction changes, are then taken as reset points. By averaging the x and y coordinates of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling.

Apart from reset points, we need a measure of illusion strength. We take a trajectory sample halfway between the starting point and the detected reset point. We then use the angle of a straight line drawn through that point and the start point as a measure of illusion strength. The average response in the line orientation task is also calculated for each of those 54 combinations of partcipants and stimulus conditions, as a verification for the measure of illusion strength taken from the trajectories.

@@

In Figure 4, we show the distributions of direction halfway to the reset points as well as trajectories for participant 4 in the condition with 3 _cps_ internal motion and 4 _s_ passes. This exemplary participant's re-traced paths show clear 'jump' resets, all of which are detected by our simple algorithm. Other participants are not that clear cut.

```{r fig4, echo=FALSE, message=FALSE, fig.width=10, fig.height=4, fig.cap="\\label{fig:fig4}**Figure 4. Illusion Strength and Example Trajectoriesin Experiment 2.** _Left_: For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show the distribution of angles for points halfway between the start point and detected resets; our measure of illusion strength. Dots below the distribution are individual participants' average illusion strengths. Gray lines indicate no illusion. _Right_: One exemplary participant's trajectories with very clear 'jump' resets at the purple dots. (Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories without resets.) Other participants' data was not this clear."}
plotExampleData()
```

@@

```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=8, fig.cap="\\label{fig:fig5}**Figure 5. Measures of Illusion Strength** Line angle responses and reset point direction should both measure illusion strength. **Left:** Line Angle response as a function of internal and external speed, averaged across 9 participants in each of the 6 conditions. Shaded areas denote 95% confidence intervals. **Right:** Angles of samples halfway to a reset over line angle response. Each dot corresponds to one participants average in one of the 6 conditions. Diagonal line indicates a weighted regression (weighted by one over the square of the line angle response to counter heteroscedasticity), and the shaded area shows a 95% confidence interval for the regression."}
plotValidity()
```

As specified above, we also take the angle of the sample on each re-tracing trajectory that is halfway between the origin and the detected reset, as our main measure of illusion strength. If both the angle half-way to reset and line angle responses are good measures of illusion strength, they should be related. As can be seen from the data in Figure 5 (right), there might be a relation between the two variables, but there is also heteroscadasticity. We counter this by doing a weighted regression, with the weights set to one over the squared line angle ($\frac{1}{\alpha^2}$). Reported line angles predicted reset point direction (slope=0.71; _t_(49)=4.54, _p_<.001; $R^2$=.282; _F_(1,49)=20.6), verifying that the two measures - while imperfect - are likely both indicators of double-drift illusion strength. From here on, we will therefore only use the angle of samples half-way to detected reset points as measure of illusion strength.
